Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-02 Manuscript/proto-manuscript.tex
Update methods section

The partial atomic charges in original Lipid14 parametrization were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum.
If the charges are obtained by using RESP in an implicit solvent, 
they vary the most for the polar moieties \cite{maciejewski14}. 
In order to implicitly capture solvent induced polarization 
in models with fixed charges, 
we shall use the average of partial charges 
from both environments (i.e. vaccum and solvent),
so called implicitly polarized charges (IPolQ) \cite{ipolq2013}. 
By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}

Hydrocarbon chains in Lipid14 and other lipids models are highly optimized and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16},
in contrast to glycerol backbone and headgroup regions which require
some improvement in all available lipid models \cite{botan15}. 
To minimize the detuning of the highly optimized hydrocarbon
parameters, we apply ECC only
to the headgroup, glycerol backbone and carbonyl in acyl chains.
These regions are also the most polar parts in lipids and are expected
to most affect cation binding.   
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications.

Mere scaling of partial charges in the modified region by the factor $f_q$ reduces the 
area per molecule to ??\todo{find the value}, which is significantly smaller than the
experimental value (\cite{}) and the original Lipid14 values (\cite{})\todo{Add the values}.  
The decrease of area was speculated 
\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
to arise from reduced hydration
of headgroup due to the lower polarity of molecules with scaled charges.
The hydration can be increased by decreasing the radius of atoms
by reducing the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
This allows water molecules to approach closer to lipid atoms and
have stonger electrostatic interactions with them.
\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Indeed, by reducing $\sigma$ with a scaling factor $f_\sigma = 0.89$ 
we increased the area per molecule
to a level close to experimental and original Lipid14 values. 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison to experimental data}
Structures sampled by individual lipid molecules in simulations were compared
to experimental data by using C-H bond order parameters \cite{ollila16}
\begin{equation}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between C-H bond and membrane
normal and average is taken over all sampled configurations.

Bilayer dimensions were compared to experiments by using the
scattering form factor \cite{ollila16}
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

Ion binding was compared between experiments and simulations by 
using lipid headgroup order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid headgroup
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}. More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid headgroup dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}.

@@

The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
The code AmberTools \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.

>>>>

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16,??},
it is not clear if the same factor should be used for lipid headgroups \todo{, because ??}.
Here we estimated the correct scaling factor for Lipid14 model by using
so called implicitly polarized charges (IPolQ) \cite{ipolq2013},
which are the average of partial charges given by RESP~\cite{RESP_paper} calculations in
vacuum and in implicit solvent.
%In order to implicitly capture solvent induced polarization 
%in models with fixed charges, 
%we shall use the average of partial charges 
%from both environments (i.e. vaccum and solvent),
%so called implicitly polarized charges (IPolQ) \cite{ipolq2013}. 
The partial charges in original Lipid14 parametrization were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum,
while charges from calculations in an implicit solvent vary most for the polar moieties \cite{maciejewski14}. 
Here we estimate the appropriate scaling factor for lipid hedgroups and glycerol backbone
by comparing IPolQ charges (average of RESP calculation in vacuum and in implicit solvent) to
the RESP results in vacuum. By using the results reported in \cite{maciejewski14} we conclude
that the IPolQ charges can be achived from vacuum calculations by scaling roughly with the
factor of $f_q=0.8$, which is the used to derive ECC-corrected Lipid14 model.
\todo{SAMULI: please check that this is correctly descrived.}
%By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
%we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
%by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}

Here we apply ECC correction with the proposed scaling factor only to the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
highly optimized in Lipid14 and other lipids models and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16},
in contrast to glycerol backbone and headgroup regions which require
some improvement in all available lipid models \cite{botan15}. 
%To minimize the detuning of the highly optimized hydrocarbon
%parameters, 
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications. 
The scaling of partial charges in the polar region was found to decrease
area per molecule to ??\todo{find the value}, which is significantly smaller than the
experimental value (\cite{}) and the original Lipid14 values (\cite{})\todo{Add the values}.  
The decrease of area was found to arise from lower hydration of the lipid headgroup region,
%\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
%JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The solvation free energy can be decreased
and hydration increased by reducing the radius of beads by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
%This allows water molecules to approach closer to lipid atoms and
%have stonger electrostatic interactions with them.
%\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Here we reduce $\sigma$ with a scaling factor of $f_\sigma = 0.89$ for the same atoms for which
charges was scaled. This increased the area per molecule back
to a level close to experimental and original Lipid14 values. 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept}
Ion binding was compared between experiments and simulations by 
using lipid headgroup order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid headgroup
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid headgroup dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for C-H bonds can be accurately measured for each
C-H bond of lipids by using H$^2$ NMR or $^{13}$C NMR techniques and
calculated from MD simulations directly from definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

The change of the headgroup order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Combination of atomic absorption spectra, $^2$H NMR experiments
and information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.

Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}. However,
it also turned out that also the sensitivity of the order parameters response
to bound charge has to be quantified more accurately for quantitative comparison
of binding affinity. In this work we first we first quantify the response of
headgroup order parameters against experimental data from cationinc surfactants
embedded in PC bilayer~\cite{scherer89}. Essentially all surfactants locate in
lipid bilayers in these experiments, thus the amount of bound charge is known
and can be plotted against order parameter change.

@@

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
\todo{Table about simulations details and some related text is deleted in commit
  https://github.com/ohsOllila/NMRlipids\_VI-NewIonModel/commit/6526079062ca322e90f8ea0654597394b023041e
  Maybe this is done accidentally, or is there some other plan with this?
}

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
The code AmberTools \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
These systems were ran with original Lipid14 POPC model and with ECC-corrected model.
\todo{Describe the changes in surfactant model used with ECC-corrected lipids.}
